Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2023-05-12 doc/manuscript/ESSDmanuscript.Rmd
edit through end of section 6.3

New data collection and analysis is needed to fill notable knowledge gaps.
While aboveground biomass stocks in particular have received -- and continue to receive -- significant research attention, other pools and variables remain poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021].
Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
For instance, C cycling of tropical forests -- particularly in Africa -- remains relatively poorly characterized, in large part due to substantial barriers to data collection and distribution [@delima_making_2022]
*(add some more here?)*

Several variables of value for IPCC C accounting have not been calculated and presented as would be possible given the same forest census data and minimal extra research effort.
For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022]. 
Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be easily be estimated in parallel with aboveground biomass, with the greatest barrier being availability of reliable allometries, as have been developed for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
However, while equations for estimating root (and thereby total) biomass require improvement, they do exist *for many forest types*[@refs], and IPCC provides default recommendations of below-ground to above-ground ratios for estimation of root biomass [@ipcc_2019_2019].
In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate accounting. 
We recommend that, when possible, researchers calculate these, following the reporting guidelines specified in section `r section_data_reporting_needs`.

*Other EFDB-relevant variables require more effort but are warranted given their importance for forest C accounting.*
Given widespread trends of increasing tree mortality [@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.
...

@@

We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards (Table `r table_recommendations`).
It is particularly noteworthy that simple decisions on the presentation of results will determine whether the data meet the criteria for inclusion in EFDB.
Some examples are as follows:
(1) presenting data only in a figure makes it ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion;
(2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies data from inclusion;
(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant for IPCC accounting (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC. Reports which had the most successful data transfers used EFDB variables and had clear tables showing their results.

@@

library(knitr)
library(kableExtra)
knitr::kable(recommendations_table, format = "latex",
             longtable = FALSE,
             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....") %>%
  row_spec(0, bold = TRUE)%>%
  kable_styling(font_size = 10) %>%
  kableExtra::kable_styling(full_width = TRUE) 
  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```

For those compiling published data (e.g., for meta-analyses), the data set can have added value if all information required by IPCC is extracted from original publications. 
This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.

Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipline developed here.
For individual publications, the former option will generally be more efficient.
However, by getting the data into ForC as well as EFDB, the latter option will allow the data to be more broadly useful--for example, being used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021] or model benchmarking [@fer_ecosystem_2021].

>>>>

New data collection and analysis is needed to fill notable knowledge gaps.
While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021], production of an accurate global map of forest C stocks remains an ongoing challenge [@refs]. 
Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@refs].
Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@refs].

Several EFDB-relevant variables have not been calculated and presented as frequently as would be possible given existing forest census data and minimal extra research effort.
For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022]. 
Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be estimated in parallel with aboveground biomass, with the greatest barrier being that allometric models for estimating root biomass are not as reliable or easily available as are those for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017].
In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate accounting. 
We recommend that, when possible, researchers calculate and report these variables, following the reporting guidelines specified in section `r section_data_reporting_needs`.

Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for forest C accounting.
Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@refs], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@refs].
Given observations of increasing tree mortality in some forested regions [@refs], including through severe natural disturbance [@refs], better characterization of forest dead wood will be critical.
Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@refs].
Significant investment in ground-based forest research will be critical to filling these gaps.

@@

We recommend that, in order to make research valuable to C accounting under IPCC guidelines, researchers calculate and report results according to IPCC standards (Table `r table_recommendations`).
It is particularly noteworthy that simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
Some examples are as follows:
(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion;
(2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;
(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant for IPCC accounting (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC. Reports which had the most successful data transfers used EFDB variables and had clear tables showing their results.

@@

library(knitr)
library(kableExtra)
knitr::kable(recommendations_table, format = "latex",
             longtable = FALSE,
             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.}") %>%
  row_spec(0, bold = TRUE)%>%
  kable_styling(font_size = 10) %>%
  kableExtra::kable_styling(full_width = TRUE) 
  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```

For those compiling published records (e.g., for meta-analyses), the data set can have added value if all information required by EFDB is extracted from original publications. 
This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.

Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipeline developed here.
For individual publications, the former option will generally be more efficient.
However, data incorporated into ForC as well as EFDB will be more broadly useful; for example, these data may be used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021], analyses of forest-based climate change mititagation potential [e.g., @cook-patton_mapping_2020; @goldstein_protecting_2020], and model benchmarking [@fer_ecosystem_2021].

### 2023-05-12 doc/manuscript/ESSDmanuscript.tex
edit through end of section 6.3

New data collection and analysis is needed to fill notable knowledge
gaps. While aboveground biomass stocks in particular have received --
and continue to receive -- significant research attention, other pools
and variables remain poorly quantified \citep[Table
2,][]{anderson-teixeira_carbon_2021}. Furthermore, data distribution is
uneven across forest types and geographical regions (Figs. 2-3). For
instance, C cycling of tropical forests -- particularly in Africa --
remains relatively poorly characterized, in large part due to
substantial barriers to data collection and distribution
\citep{delima_making_2022} \emph{(add some more here?)}

Several variables of value for IPCC C accounting have not been
calculated and presented as would be possible given the same forest
census data and minimal extra research effort. For example, aboveground
woody mortality (\emph{woody.mortality\_ag}) and aboveground biomass
increment (\emph{delta.agb}) can be calculated from the same census data
as aboveground woody productivity (\emph{ANPP\_woody}), yet the latter
has received far more research attention, and correspondingly has far
more records in ForC \citetext{\citealp[Table
2,][]{anderson-teixeira_carbon_2021}; \citealp[but
see][]{piponiot_distribution_2022}}. Similarly, live coarse root
biomass, total biomass, and changes in both of these pools could in
theory be easily be estimated in parallel with aboveground biomass, with
the greatest barrier being availability of reliable allometries, as have
been developed for aboveground biomass
\citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022}.
However, while equations for estimating root (and thereby total) biomass
require improvement, they do exist \emph{for many forest
types}\citep{refs}, and IPCC provides default recommendations of
below-ground to above-ground ratios for estimation of root biomass
\citep{ipcc_2019_2019}. In addition, standing dead trees are captured in
most forest censuses and could be used to estimate standing dead wood,
although additional data on breakage would be needed for accurate
accounting. We recommend that, when possible, researchers calculate
these, following the reporting guidelines specified in section 6.3.

\emph{Other EFDB-relevant variables require more effort but are
warranted given their importance for forest C accounting.} Given
widespread trends of increasing tree mortality \citep{refs}, including
through severe natural disturbance \citep{refs}, better characterization
of dead wood will be critical. \ldots{}

@@

We recommend that, unless they have some specific reason to do
otherwise, researchers calculate and report the values according to IPCC
standards (Table 3). It is particularly noteworthy that simple decisions
on the presentation of results will determine whether the data meet the
criteria for inclusion in EFDB. Some examples are as follows: (1)
presenting data only in a figure makes it ineligible for inclusion in
EFDB, whereas presentation in a table or supplementary data file allows
inclusion; (2) direct presentation of all relevant variables allows
inclusion, whereas presenting only components of variables of interest
(e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or
requiring simple mathematical operations to obtain a variable of
interest (e.g., \emph{delta.agb} = \emph{ANPP\_woody} -
\emph{woody.mortality.agb}) disqualifies data from inclusion; (3)
matching IPCC-defined thresholds for defining C pools (Table 1), which
may vary by country, can make the data far more relevant for IPCC
accounting (e.g., using a 10 cm cutoff between dead wood and litter,
presenting soil C to a depth of 30 cm). It should also be emphasized
that reporting of 95\% confidence intervals (or other metrics of error),
when applicable, is highly desirable and makes the data more relevant to
IPCC. Reports which had the most successful data transfers used EFDB
variables and had clear tables showing their results.

@@

\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....}
\centering
\fontsize{10}{12}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{criteria} & \textbf{recommendation} & \textbf{rationale}\\
\hline
variables to include & When possible, calculate and present all relevant variables that can be readily caluclated based on available data. & Estimates of relevant variables are not always calcualted.\\
\hline
forest census methods & Adopt IPCC guidelines (country-specific) for minimum stem size in censues in census and reporting. Ideally, census stem down to the smallest diameter feasible. & IPCC biomass pool definition includes all living vegetation, but understory may be excluded when contribution is minor.\\
\hline
 & *taxa to include* & \\
\hline
dead organic matter sampling & *include damage estimates on standing dead trees* & \\
\hline
 & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & \\
\hline
belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & IPCC biomass pool definition includes all living vegetation\\
\hline
reporting variiables & Present each variable individually, as opposed to requiring that variables of interest be calculated from related variables. & EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
\hline
reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
\hline
reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals, when possible.\\
\hline
\end{tabu}
\end{table}

For those compiling published data (e.g., for meta-analyses), the data
set can have added value if all information required by IPCC is
extracted from original publications. This includes -- but is not
limited to -- retaining original values as presented without
modification or rounding, noting whether data were digitized, recording
confidence intervals, and recording all required fields (as indicated in
the EFDB's bulk import template). The significant effort required to map
a database into EFDB has been accomplished here (Appendix B), and we
welcome other researchers to use the ForC template.

Once EFDB-relevant data are available in peer-reviewed publications,
they may be submitted directly to EFDB or may use the ForC - EFDB data
pipline developed here. For individual publications, the former option
will generally be more efficient. However, by getting the data into ForC
as well as EFDB, the latter option will allow the data to be more
broadly useful--for example, being used for basic science
\citep[e.g.,][]{banburymorgan_global_2021, anderson-teixeira_carbon_2021}
or model benchmarking \citep{fer_ecosystem_2021}.

>>>>

New data collection and analysis is needed to fill notable knowledge
gaps. While aboveground biomass stocks in particular have received --
and continue to receive -- by far the most research attention
\citep[Table 2,][]{anderson-teixeira_carbon_2021}, production of an
accurate global map of forest C stocks remains an ongoing challenge
\citep{refs}. Other pools and variables remain very poorly quantified
\citep[Table 2,][]{anderson-teixeira_carbon_2021}, introducing
substantive uncertainties into global forest C budgets \citep{refs}.
Furthermore, data distribution is uneven across forest types and
geographical regions (Figs. 2-3). For instance, data on C cycling of
tropical forests -- particularly in Africa -- remains relatively sparse,
in large part due to substantial barriers to data collection and
distribution \citep{delima_making_2022}. Significant investment in
research and researchers focused on ground-based measurement of forest C
in such regions will be important to filling knowledge gaps in forest C
cycling \citep{delima_making_2022, refs}.

Several EFDB-relevant variables have not been calculated and presented
as frequently as would be possible given existing forest census data and
minimal extra research effort. For example, aboveground woody mortality
(\emph{woody.mortality\_ag}) and aboveground biomass increment
(\emph{delta.agb}) can be calculated from the same census data as
aboveground woody productivity (\emph{ANPP\_woody}), yet the latter has
received far more research attention, and correspondingly has far more
records in ForC \citetext{\citealp[Table
2,][]{anderson-teixeira_carbon_2021}; \citealp[but
see][]{piponiot_distribution_2022}}. Similarly, live coarse root
biomass, total biomass, and changes in both of these pools could in
theory be estimated in parallel with aboveground biomass, with the
greatest barrier being that allometric models for estimating root
biomass are not as reliable or easily available as are those for
aboveground biomass
\citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022}.
However, while equations for estimating root (and thereby total) biomass
require improvement, they do exist for many forest types
\citep[e.g.,][]{brassard_coarse_2011, chojnacky_updated_2014, waring_overlooking_2017}.
In addition, standing dead trees are captured in most forest censuses
and could be used to estimate standing dead wood, although additional
data on breakage would be needed for accurate accounting. We recommend
that, when possible, researchers calculate and report these variables,
following the reporting guidelines specified in section 6.3.

Filling knowledge gaps in other EFDB-relevant variables will require
more effort, but this effort is warranted given their importance for
forest C accounting. Although aboveground biomass is the most studied
variable considered here (Table 2) and is the target of satellite
missions \citep{refs}, significant ground-based research effort is
required to create accurate global maps of forest biomass and changes
therein \citep{refs}. Given observations of increasing tree mortality in
some forested regions \citep{refs}, including through severe natural
disturbance \citep{refs}, better characterization of forest dead wood
will be critical. Additionally, C stocks in forest organic horizons and
soils can be quite substantial and highly uncertain in many parts of the
world \citep{refs}. Significant investment in ground-based forest
research will be critical to filling these gaps.

@@

We recommend that, in order to make research valuable to C accounting
under IPCC guidelines, researchers calculate and report results
according to IPCC standards (Table 3). It is particularly noteworthy
that simple decisions on the presentation of results will determine
whether the records meet the criteria for inclusion in EFDB. Some
examples are as follows: (1) presenting data only in a figure makes them
ineligible for inclusion in EFDB, whereas presentation in a table or
supplementary data file allows inclusion; (2) direct presentation of all
relevant variables allows inclusion, whereas presenting only components
of variables of interest (e.g., parsing litter into fine woody debris,
OL, OF, and OH layers) or requiring simple mathematical operations to
obtain a variable of interest (e.g., \emph{delta.agb} =
\emph{ANPP\_woody} - \emph{woody.mortality.agb}) disqualifies records
from inclusion; (3) matching IPCC-defined thresholds for defining C
pools (Table 1), which may vary by country, can make the data far more
relevant for IPCC accounting (e.g., using a 10 cm cutoff between dead
wood and litter, presenting soil C to a depth of 30 cm). It should also
be emphasized that reporting of 95\% confidence intervals (or other
metrics of error), when applicable, is highly desirable and makes the
data more relevant to IPCC. Reports which had the most successful data
transfers used EFDB variables and had clear tables showing their
results.

@@

\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.}}
\centering
\fontsize{10}{12}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{criteria} & \textbf{recommendation} & \textbf{rationale}\\
\hline
variables to include & When possible, calculate and present all relevant variables that can be readily estimated based on available data. & Estimates of relevant variables are not always calcualted.\\
\hline
forest census methods & Adopt IPCC guidelines (country-specific) for minimum stem size in censues in census and reporting. Ideally, census stem down to the smallest diameter feasible. & IPCC biomass pool definition includes all living vegetation, but understory may be excluded when contribution is minor.\\
\hline
 & Census all taxa crontributing signficantly to biomass & IPCC biomass pool definition includes all living vegetation.\\
\hline
dead organic matter sampling & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & Diameter cutoff must be applied consistently by each country.\\
\hline
belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & Diameter cutoff must be applied consistently by each country.\\
\hline
reporting variables & Present each EFDB- relevant variable individually, as opposed to requiring summation of related variables. & EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
\hline
reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
\hline
reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals whenever possible.\\
\hline
\end{tabu}
\end{table}

For those compiling published records (e.g., for meta-analyses), the
data set can have added value if all information required by EFDB is
extracted from original publications. This includes -- but is not
limited to -- retaining original values as presented without
modification or rounding, noting whether data were digitized, recording
confidence intervals, and recording all required fields (as indicated in
the EFDB's bulk import template). The significant effort required to map
a database into EFDB has been accomplished here (Appendix B), and we
welcome other researchers to use the ForC template.

Once EFDB-relevant data are available in peer-reviewed publications,
they may be submitted directly to EFDB or may use the ForC - EFDB data
pipeline developed here. For individual publications, the former option
will generally be more efficient. However, data incorporated into ForC
as well as EFDB will be more broadly useful; for example, these data may
be used for basic science
\citep[e.g.,][]{banburymorgan_global_2021, anderson-teixeira_carbon_2021},
analyses of forest-based climate change mititagation potential
\citep[e.g.,][]{cook-patton_mapping_2020, goldstein_protecting_2020},
and model benchmarking \citep{fer_ecosystem_2021}.
